A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-04-25 jacobian_calc_CMAME.tex
added displacement label to truss figure
L_0 labeled in the truss figure
Final major edits

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Newton's method \sep Newton-Raphson \sep numerical differentiation \sep complex-step \sep finite difference \sep automatic differentiation \sep Jacobian \sep tangent-stiffness
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%\MSC [2010] 65D25 
\end{keyword}

@@

A goal of this study was to develop and evaluate a \todo[]{Re: I.1} relatively unexplored, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This  \todo[]{Re: I.1} method is based on a complex number Taylor series expansion and referred to as CTSE or the ``complex-step'' method. The practical accuracy of CTSE is determined by comparing its measured accuracy relative to the measured accuracy of popular finite differencing techniques used for computing tangent-stiffness matrices. For this comparison the,  exact to machine precision, derivatives computed with \emph{automatic differentiation} serve as the standard.

@@

\todo[]{Re: I.4}As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler \cite{lyness1967numerical}.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers \cite{lyness1968differentiation}.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through numerical experiments that CS offers superior accuracy to FD and CD \cite{squire1998using} in computing numerical derivatives. Papers by Newman et al.\ \cite{newman1998} and Anderson et al.\ \cite{anderson2001sensitivity} discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute Jacobians of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ \cite{perez2000numerical, perez2012numerical} used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ \cite{martins2000automated} investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.

Martins et al.\ \cite{martins2003complex} demonstrates that the CS method can be performed by using complex datatypes for the variables of functions comprising an analysis code. The accuracy of this approach, with respect to the derivatives themselves, is compared to automatic-differentiation in a serial analysis code. Lai and Crassidis \cite{lai2008extensions} discuss first- and second-order CS methods for computing derivatives and choosing optimal step sizes.  Al-Mohy and Higham \cite{al2010complex} investigate the use of CS in computing Fr\'{e}chet derivatives of matrix functions. Jin et al.\ \cite{jin2010improved} discuss using CS to compute sensitivities in a solid mechanics FEM analysis.  Voorhees et al.\ \cite{voorhees2011complex} discusses the background of the CS method and frames it as a subset of Fourier Differentiation. CS and Fourier Differentiation are then applied to computing sensitivities within a solid mechanics finite element analysis. They implemented the complex-values as additional degrees of freedom in the finite element stiffness matrices using a real-valued matrix representation of the complex-numbers where required. In a follow-on work,  Millwater et al.\  \cite{millwater2013application} extended this concept, calling it ZFEM, and implemented the method into user defined element types for use with the popular ABAQUS \cite{systemes2012abaqus} FEA software package.  Nishida et al.\ \cite{nishida2013} used the complex-step method to compute Jacobian matrices for implementation in a Levenberg-Marquardt optimization routine for design-of-experiments purposes.  Finally, Abreu et al.\ \cite{abreu2013generalization} discusses computing derivatives with a version of the CS method which features the combined use of real and imaginary valued step components, as in Fourier differentiation.  They suggests that the modified CS method can achieve extended approximation accuracy up to fourth order.  This recent literature has introduced the application of CS to single-dimension and multidimensional derivatives, sensitivity analysis, non-linear solution schemes, optimization, the extension of CS for higher accuracy, and implementation strategies for CS in new and existing codes.  The current paper presents a comparison of the complex-step method to the other methods for numerical differentiation in as close to a one-to-one implementation as possible for direct comparison purposes in a large-scale engineering analysis code. To the author's knowledge it is the first application of the complex-step method to solve peridynamics mechanics problems, and the first implementation and comparison of the method in large-scale distributed memory parallel setting.

@@

The example problem is a single degree-of-freedom nonlinear mechanical system involving a connected truss member and spring as illustrated in Fig. \ref{fig:TrussSchematic}. 
%
\begin{figure}[tbp] 
\centering %%\includegraphics[width=0.7\textwidth]{./figs/truss.png}

@@

@@

@@

@@

@@

@@

@@

\todo[]{Re: I.2} Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic} is a nonlocal reformulation of the partial-differential equations that provide the statement of momentum balance in classical continuum mechanics. Its primary goal is to avoid the use of spatial derivatives in the balance equations or constitutive laws such that discontinuous displacements, i.e. cracks, are mathematically consistent with the governing equations. It has demonstrated great promise in modeling problems with pervasive failure \cite{littlewood2010}, interesting phenomenon such as dynamic crack branching \cite{ha2010sod}, and is being extended to multiphyics phenomena \cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale coupling to molecular simulations \cite{seleson2009peridynamics,seleson2014atom,rahman2013b}.  The material model used in this study was a linear peridynamic solid which replicates linear elasticity in the \emph{local limit}, i.e.\ when the non-local length scale vanishes. While \emph{linear} may appear in the name of the material model, it is geometrically non-linear under large deformations and requires a nonlinear solver to converge to equilibrium conditions in these situations.  The peridynamic equation of motion is shown in equation~(\ref{eqn:PDMotion})
%
\begin{equation} 
    \rho(\mathbf{x})\mathbf{\ddot{u}(x)} = \int_{\mathcal{H_{\delta(\mathbf{x})}}} \mathbf{f(x,x'}) {\rm d}V_{\mathbf{x'}} + \mathbf{b(x)},
    \label{eqn:PDMotion}
\end{equation}

where, $\rho$ is the mass density, $\mathbf{\ddot{u}}$ is the second time derivative of displacement (i.e.\ acceleration), $\mathcal{H}_{\delta(\mathbf{x})}$ is a neighborhood containing remote material points, $\mathbf{x'}$, with characteristic length $\delta$, centered at $\mathbf{x}$. The integrand $\mathbf{f}$ contains the constitutive response of the material.  Physically, $\delta$ is a material length-scale and computationally it, along with the discretization, will set the bandwidth of the tangent-stiffness matrix. $\mathbf{b}$ is a body force density.  An explicit dependence on time $t$ has been supressed from the functional arguments, but is implied.  The reader is refered to the seminal work on peridynamics \cite{silling:psa} for detailed derivations of the momentum equation and example consitutive models.

@@

\subsubsection{Data reduction} 
The output produced by \emph{Peridigm} including the additional accuracy and speed measurements was redirected from the console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).  Directories and filenames were chosen so that data would be easy to sort by individual test run.  \emph{Python} scripts were then used to post-process and plot the data. These scripts averaged accuracy or speed data for a tests corresponding to a particular peridynamic node density over all iterations for that run, and then plotted this averaged data  as a function of peridynamic node density as it varied from test to test in the series.  Similarly, for the parallel tests, the data was averaged and plotted in much the same way, but as a function of number of compute cores used to solve the problem. Confidence intervals were not calculated on the measurements. The main reason is that network traffic from other users makes running tests on the HPC cluster a time varying process. These data reduction and plotting scripts can be found in the paper repository.

@@

\subsection{Speed data} 
%
Speed measurements were taken according to the methods mentioned in Section~\ref{tcs}. The averaged results of these speed measurements for the serial tests can be seen in Fig.~\ref{fig:serial_speed}.  Average time in seconds to compute a TS matrix is plotted as a function of the number of nonzero TS elements.  CS was the slowest of the four methods, taking the longest average time per iteration, followed by CD, FD and AD in that order. Calculation time has a power-law relationship with problem size, with a power-law index of roughly $\num{1.0E-6}$ for each of the methods. This bodes well for the scalability of each of the methods by themselves in \emph{Peridigm}. If we assumed every individual operation, generally speaking, took the same amount of time on average such that compute time were a measure of number of operations, this suggests that the methods as in \emph{Peridigm} have the same complexity class when considering computing a TS matrix alone, that is computation time is $\mathcal{O}(10^{(10^{-6} \times \log_{10}n)})$. The definition of complexity class used here is the informal one given in \cite[pp. 1059]{cormen2009introduction}. This would suggest that the methods are equally well suited to the task of computing a TS matrix, if it were further assumed that the TS matrices produced by each method were identical. This is not the case as seen in the following subsection.
%
\begin{figure}[tbp] 
    \centering 
    \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
    \caption{Serial test series speed measurements.} 
    \label{fig:serial_speed}
\end{figure}
%
For the parallel simulations, averaged speed results appear in Figure~\ref{fig:multi_speed}. For this series, rather than varying the number of nonzero TS matrix elements, the number of compute cores used to solve the problem is increased, while the number of nonzero TS matrix elements remained constant at $\num{1.67E9}$ .  CS again was the slowest method, followed by CD, AD and then FD.  There is a somewhat surprise reversal of FD and AD for these parallel simulations.  Figure~\ref{fig:multi_speed} indicates that when looking at average iteration time, the factor of speed improvement scales inversely with the number of compute cores used for the given conditions of this group of tests. Figure~\ref{fig:multi_speed} clearly shows that when core count doubles, compute time halves, for each method. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale well when additional processors are used. 
%
\begin{figure}[tbp] 
    \centering 
    \scalebox{1.0}{\input{./figs/multi_speed.pgf}}
    \caption{Multicore test series speed measurements.} 
    \label{fig:multi_speed}
\end{figure}
%
A speculative explanation as to why FD and AD have reversed places in the speed tests with respect to the serial results is that the AD data-structures carry a list of partial derivatives of the functional evaluations along side the functional evaluations themselves only to be processed at the end of the tangent-stiffness matrix calculation when the derivatives are requested. This represents additional load on the hardware memory subsystem, such that may exceed that of FD yet still falls under a supposed memory bandwidth threshold during serial operation. Cores within a multicore processor use the same shared bus to access memory, such that cores compete for memory access. Supposing that exceeding the memory bandwidth threshold penalizes additional requested bandwidth, it is likely that the decrease in available memory bandwidth per core caused by parallel operation coupled with AD's additional resource requirements causes the AD method to be slower than FD.

@@

    \scalebox{1.0}{\input{./figs/multi_accuracy.pgf}} 

@@

\subsection{Efficiency data}
%
Efficiency, calculated as the average number of seconds taken per nonzero Jacobian matrix element, was plotted as a function of the number of nonzero TS matrix elements. The single core results are shown in Figure~\ref{fig:serial_efficiency}.  The ranking order of the methods matched their speed ranking in Figure~\ref{fig:serial_speed}. This plot indicates that efficiency is insensitive to the number of nonzero TS elements.  
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed_rel.pgf}}
  \caption{Serial test series efficiency measurements.}
  \label{fig:serial_efficiency}
\end{figure}
%
The parallel efficiency results are shown in Figure~\ref{fig:multi_efficiency}. The parallel efficiency plot shows that AD was slower than FD because for the given test problem it had a lower efficiency. Strangely AD appears to gain in efficiency over the last few data points in the plot. The reason behind this trend was not determined.
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_speed_rel.pgf}}
  \caption{Multicore test series efficiency measurements.}
  \label{fig:multi_efficiency}
\end{figure}

\section{\todo[]{Re: I.5}Convergence and displacement accuracy studies.} 
\label{sec:PeridigmConvergenceStudy}
%
While the previous sections profiled the per-iteration performance of each of the methods, it was necessary to conduct convergence rate studies in order to get a complete picture. It is necessary for an analyst to know if the accuracy differences measured in the per-iteration tests would amount to differences in convergence rates among the methods if they were allowed to solve problems normally. The question is posed, given that step-size $h$ is heuristically selected by \emph{Peridigm} for the finite difference methods and presumable makes a good choice, do the single-iteration Jacobian accuracy results predict the ranking of the methods in terms of convergence rate for a given study? For brevity, convergence rate was measured for each of the methods in solving just the largest problem of the single-core test series. That problem was solved with each of the methods both on a single core and all sixteen cores of a 'normal' node on \emph{Stampede}. As was seen in the per-iteration results, Jacobian accuracy did not vary greatly from test to test, such that any pair of tests from the two series could have been considered reasonably representative. As in the validation problem shown in Section~\ref{subsec:Validation}, would each of the methods return an equal convergence rate so long as $h$ were optimally selected?

The results shown in Table~\ref{tab:ConvergenceStudy2} indicate that this was the case.  The difference in Jacobian accuracy between the methods seen in the per-iteration tests was not sufficient to lead to differences in the convergence rate of the methods.
%
\begin{table}[htb]   
    \centering 
    \caption{Average iterations / load-step} 
    \label{tab:ConvergenceStudy2}   
    \begin{tabular}{c c c c c}
        \toprule 
        Test type & AD & CS & CD & FD\\
        \midrule 
        single-core & 63 & " & " & "\\ 
        multi-core  & 63 & " & " & "\\ 
        \bottomrule 
    \end{tabular} 
\end{table}

The end displacement results shown in Section~\ref{subsec:Validation} indicated that up to a certain precision the solutions produced by each of the methods were identical. It was examined if this was the case for the solutions produced in \emph{Peridigm}. Here each of the solutions produced by each of the methods were compared to the AD produced solutions, under the assumption that the AD produced solutions would be closest to the solution produced by using an 'exact' Jacobian. Table~\ref{tab:PeridigmSolutionAccuracy} shows that the methods did indeed produce differing solutions in terms of nodal force density, although only in decimal places at least $9$ orders of magnitude smaller than the force density value at the node where the difference was detected.
%
\begin{table}[hbp]   
    \centering 
    \caption{Nodal force density maximum difference} 
    \label{tab:PeridigmSolutionAccuracy}   
    \begin{tabular}{c c c c}
        \toprule  & CS & CD & FD\\
        \midrule  Order of magnitude & $\num{1.E-8}$ & $\num{1.E-8}$ & $\num{1.E-5}$ \\ 
        \bottomrule 
    \end{tabular} 
\end{table}

The solution accuracy results do follow the same ranking seen in the Jacobian accuracy results; however, again as in the validation problem the results differ only slightly. Speculatively this degree of error may be important to those running probabilistic simulations such that the propagation of error may be a concern. Also it is conceivable that if this error is a fixed quantity, insensitive to material properties or loading conditions, it may have a greater impact on simulations involving very low force densities such as for small displacements or very compliant materials. In this case it could be recommended that forward difference be avoided. 
%
\section{\todo[]{Re: I.5}Conclusions}
\label{sec:Conc}
%
\subsection{For further studies of Jacobian accuracy in Newton's method}
\label{subsec:FurtherStudies}
The question is raised; what is the significance of Jacobian estimation accuracy to the convergence of Newton's method?  The per-iteration and even convergence studies shown here indicate the effect of Jacobian accuracy for only a single datum, a very accurate Jacobian estimate, while the validation problem of Section~indicates the effect of using a grossly inaccurate Jacobian. The results here show that for very small Jacobian estimate inaccuracies, convergence rate is not affected, while for grossly inaccurate Jacobians Newton's method fails.  Additional confounding factors include that a Jacobian is only first order approximation of an often curved energy surface and that a non-naive implementation of Newton's method includes a line search phase that helps in choosing an optimal iterate given a Jacobian.  Is it worthwhile then to investigate the effect of using moderately inaccurate Jacobians, produced deliberately or otherwise, on the convergence rate of Newton's method?

The answer may be no at least unless a more highly non-linear class of problem is selected for study or a controllable method of producing a moderately inaccurate Jacobian is found along with a reason for using it. While it is possible to incorrectly apply finite differencing and CS by choosing an inappropriate step-size, there is no reason that this should be done. The experiments show that step-size $h$ can be properly selected and that the methods are capable of producing accurate Jacobians. It should be noted that the model used to perform the simulation could always be substituted for a more highly non-linear one and perhaps this would somewhat increase the importance of Jacobian accuracy. The study presented here failed to pose the question of how wrong a correctly implemented Jacobian calculation scheme could go for plastic or otherwise non-linear material models, here instead non-linearity was imposed by the nonlocal interaction between material points essential to peridynamic models, like the linear model used in this study.

>>>>

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Newton's method \sep Newton-Raphson \sep numerical differentiation \sep complex-step \sep finite difference \sep automatic differentiation \sep tangent-stiffness \sep tangent-stiffness
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%\MSC [2010] 65D25 
\end{keyword}

@@

A goal of this study was to develop and evaluate a \todo[]{Re: I.1} relatively unexplored, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This  \todo[]{Re: I.1} method is based on a complex number Taylor series expansion and referred to as the \emph{complex-step} (CS) method. The practical accuracy of CS is determined by comparing its measured accuracy relative to the measured accuracy of popular finite differencing techniques used for computing tangent-stiffness matrices. For this comparison the,  exact to machine precision, derivatives computed with \emph{automatic differentiation} serve as the standard.

@@

\todo[]{Re: I.4}As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler \cite{lyness1967numerical}.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers \cite{lyness1968differentiation}.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through numerical experiments that CS offers superior accuracy to FD and CD \cite{squire1998using} in computing numerical derivatives. Papers by Newman et al.\ \cite{newman1998} and Anderson et al.\ \cite{anderson2001sensitivity} discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute tangent-stiffnesss of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ \cite{perez2000numerical, perez2012numerical} used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ \cite{martins2000automated} investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.

Martins et al.\ \cite{martins2003complex} demonstrates that the CS method can be performed by using complex datatypes for the variables of functions comprising an analysis code. The accuracy of this approach, with respect to the derivatives themselves, is compared to automatic-differentiation in a serial analysis code. Lai and Crassidis \cite{lai2008extensions} discuss first- and second-order CS methods for computing derivatives and choosing optimal step sizes.  Al-Mohy and Higham \cite{al2010complex} investigate the use of CS in computing Fr\'{e}chet derivatives of matrix functions. Jin et al.\ \cite{jin2010improved} discuss using CS to compute sensitivities in a solid mechanics FEM analysis.  Voorhees et al.\ \cite{voorhees2011complex} discusses the background of the CS method and frames it as a subset of Fourier Differentiation. CS and Fourier Differentiation are then applied to computing sensitivities within a solid mechanics finite element analysis. They implemented the complex-values as additional degrees of freedom in the finite element stiffness matrices using a real-valued matrix representation of the complex-numbers where required. In a follow-on work,  Millwater et al.\  \cite{millwater2013application} extended this concept, calling it ZFEM, and implemented the method into user defined element types for use with the popular ABAQUS \cite{systemes2012abaqus} FEA software package.  Nishida et al.\ \cite{nishida2013} used the complex-step method to compute tangent-stiffness matrices for implementation in a Levenberg-Marquardt optimization routine for design-of-experiments purposes.  Finally, Abreu et al.\ \cite{abreu2013generalization} discusses computing derivatives with a version of the CS method which features the combined use of real and imaginary valued step components, as in Fourier differentiation.  They suggests that the modified CS method can achieve extended approximation accuracy up to fourth order.  This recent literature has introduced the application of CS to single-dimension and multidimensional derivatives, sensitivity analysis, non-linear solution schemes, optimization, the extension of CS for higher accuracy, and implementation strategies for CS in new and existing codes.  The current paper presents a comparison of the complex-step method to the other methods for numerical differentiation in as close to a one-to-one implementation as possible for direct comparison purposes in a large-scale engineering analysis code. To the author's knowledge it is the first application of the complex-step method to solve peridynamics mechanics problems, and the first implementation and comparison of the method in large-scale distributed memory parallel setting.

@@

The example problem is a single degree-of-freedom nonlinear mechanical system involving a connected truss member and spring as illustrated in Fig. \ref{fig:TrussSchematic}. 
\centering %%\includegraphics[width=0.7\textwidth]{./figs/truss.png}

@@

\tikzstyle{displacement_reference}=[fill,pattern=north east lines,draw=none,minimum width=5.cm, minimum height=.6cm];

@@

\coordinate (tm) at (2.5, 2.0);

@@

\coordinate (dispref) at (6.5, 3.06);
\coordinate (bmdisref) at (6.5, 3.0);
\coordinate (bldisref) at (6.0,3.0);
\coordinate (brdisref) at (7.0,3.0);
\coordinate (bbdisref) at (6.5,2.5);
\coordinate (bbrdisref) at (6.8,2.6);
\node (displacement_reference) at (dispref) [rectangle, draw, style=displacement_reference, scale=.2]{};

@@

\node (bar_length_label) at (tm) [] {$L_0$};

@@

\node (displacement_label) at (bbrdisref) [] {$D$};

@@

\path (bldisref) -- (brdisref) [draw];
\path (bmdisref) -- (bbdisref) [draw, -latex];

@@

\todo[]{Re: I.2} Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic} is a nonlocal reformulation of the partial-differential equations that provide the statement of momentum balance in classical continuum mechanics. Its primary goal is to avoid the use of spatial derivatives in the balance equations or constitutive laws such that discontinuous displacements, i.e. cracks, are mathematically consistent with the governing equations. It has demonstrated great promise in modeling problems with pervasive failure \cite{littlewood2010}, interesting phenomenon such as dynamic crack branching \cite{ha2010sod}, and is being extended to multiphyics phenomena \cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale coupling to molecular simulations \cite{seleson2009peridynamics,seleson2014atom,rahman2013b}.  The material model used in this study was a linear peridynamic solid which replicates linear elasticity in the \emph{local limit}, i.e.\ when the non-local length scale vanishes. While \emph{linear} may appear in the name of the material model, it is geometrically non-linear under large deformations and requires a nonlinear solver to converge to equilibrium conditions in these situations.  The peridynamic equation of motion is shown in equation~(\ref{eqn:PDMotion})
%
\begin{equation} 
    \rho(\mathbf{x})\mathbf{\ddot{u}(x)} = \int_{\mathcal{H_{\delta(\mathbf{x})}}} \mathbf{f(x,x'}) {\rm d}V_{\mathbf{x'}} + \mathbf{b(x)},
    \label{eqn:PDMotion}
\end{equation} 
%
where, $\rho$ is the mass density, $\mathbf{\ddot{u}}$ is the second time derivative of displacement (i.e.\ acceleration), $\mathcal{H}_{\delta(\mathbf{x})}$ is a neighborhood containing remote material points, $\mathbf{x'}$, with characteristic length $\delta$, centered at $\mathbf{x}$. The integrand $\mathbf{f}$ contains the constitutive response of the material.  Physically, $\delta$ is a material length-scale and computationally it, along with the discretization, will set the bandwidth of the tangent-stiffness matrix. $\mathbf{b}$ is a body force density.  An explicit dependence on time $t$ has been supressed from the functional arguments, but is implied.  The reader is refered to the seminal work on peridynamics \cite{silling:psa} for detailed derivations of the momentum equation and example consitutive models.

@@

\subsubsection{Data reduction} 
%
The output produced by \emph{Peridigm} including the additional accuracy and speed measurements was redirected from the console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).  Directories and filenames were chosen so that data would be easy to sort by individual test run.  \emph{Python} scripts were then used to post-process and plot the data. These scripts averaged accuracy or speed data for a tests corresponding to a particular peridynamic node density over all iterations for that run, and then plotted this averaged data  as a function of peridynamic node density as it varied from test to test in the series.  Similarly, for the parallel tests, the data was averaged and plotted in much the same way, but as a function of number of compute cores used to solve the problem. Confidence intervals were not calculated on the measurements. The main reason is that network traffic from other users makes running tests on the HPC cluster a time varying process. These data reduction and plotting scripts can be found in the paper repository.

@@

\subsection{Speed data} 
%
Speed measurements were taken according to the methods mentioned in Section~\ref{tcs}. The averaged results of these speed measurements for the serial tests can be seen in Fig.~\ref{fig:serial_speed}.  Average time in seconds to compute a TS matrix is plotted as a function of the number of nonzero TS elements.  CS was the slowest of the four methods, taking the longest average time per iteration, followed by CD, FD and AD in that order. Calculation time has a power-law relationship with problem size, with a power-law index of roughly $\num{1.0E-6}$ for each of the methods. This bodes well for the scalability of each of the methods by themselves in \emph{Peridigm}. If we assumed every individual operation, generally speaking, took the same amount of time on average such that compute time were a measure of number of operations, this suggests that the methods as in \emph{Peridigm} have the same complexity class when considering computing a TS matrix alone, that is computation time is $\mathcal{O}(10^{(10^{-6} \times \log_{10}n)})$. The definition of complexity class used here is the informal one given in \cite[pp. 1059]{cormen2009introduction}. This would suggest that the methods are equally well suited to the task of computing a TS matrix, if it were further assumed that the TS matrices produced by each method were identical. This is not the case as seen in the following subsection.
%
\begin{figure}[tbp] 
    \centering 
    \scalebox{0.93}{\input{./figs/serial_speed.pgf}}
    \caption{Serial test series speed measurements.} 
    \label{fig:serial_speed}
\end{figure}
%
For the parallel simulations, averaged speed results appear in Figure~\ref{fig:multi_speed}. For this series, rather than varying the number of nonzero TS matrix elements, the number of compute cores used to solve the problem is increased, while the number of nonzero TS matrix elements remained constant at $\num{1.67E9}$ .  CS again was the slowest method, followed by CD, AD and then FD.  There is a somewhat surprise reversal of FD and AD for these parallel simulations.  Figure~\ref{fig:multi_speed} indicates that when looking at average iteration time, the factor of speed improvement scales inversely with the number of compute cores used for the given conditions of this group of tests. Figure~\ref{fig:multi_speed} clearly shows that when core count doubles, compute time halves, for each method. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale well when additional processors are used. 
%
\begin{figure}[tbp] 
    \centering 
    \scalebox{1.1}{\input{./figs/multi_speed.pgf}}
    \caption{Multicore test series speed measurements.} 
    \label{fig:multi_speed}
\end{figure}
%
A speculative explanation as to why FD and AD have reversed places in the speed tests with respect to the serial results is that the AD data-structures carry a list of partial derivatives of the functional evaluations along side the functional evaluations themselves only to be processed at the end of the tangent-stiffness matrix calculation when the derivatives are requested. This represents additional load on the hardware memory subsystem, such that may exceed that of FD yet still falls under a supposed memory bandwidth threshold during serial operation. Cores within a multicore processor use the same shared bus to access memory, such that cores compete for memory access. Supposing that exceeding the memory bandwidth threshold penalizes additional requested bandwidth, it is likely that the decrease in available memory bandwidth per core caused by parallel operation coupled with AD's additional resource requirements causes the AD method to be slower than FD.

@@

    \scalebox{0.9}{\input{./figs/multi_accuracy.pgf}} 

@@

\subsection{Efficiency data}
%
Efficiency, calculated as the average number of seconds taken per nonzero tangent-stiffness matrix element, was plotted as a function of the number of nonzero TS matrix elements. The single core results are shown in Figure~\ref{fig:serial_efficiency}.  The ranking order of the methods matched their speed ranking in Figure~\ref{fig:serial_speed}. This plot indicates that efficiency is insensitive to the number of nonzero TS elements.  
%
\begin{figure}[tbp]
  \centering
  \scalebox{0.9}{\input{./figs/serial_speed_rel.pgf}}
  \caption{Serial test series efficiency measurements.}
  \label{fig:serial_efficiency}
\end{figure}
%
The parallel efficiency results are shown in Figure~\ref{fig:multi_efficiency}. The parallel efficiency plot shows that AD was slower than FD because for the given test problem it had a lower efficiency. Strangely AD appears to gain in efficiency over the last few data points in the plot. The reason behind this trend was not determined.
%
\begin{figure}[tbp]
  \centering
  \scalebox{0.9}{\input{./figs/multi_speed_rel.pgf}}
  \caption{Multicore test series efficiency measurements.}
  \label{fig:multi_efficiency}
\end{figure}

\section{\todo[]{Re: I.5}Convergence and displacement accuracy studies.} 
\label{sec:PeridigmConvergenceStudy}
%
While the previous sections profiled the per-iteration performance of each of the methods, it was necessary to conduct convergence rate studies in order to get a complete picture. It is necessary for an analyst to know if the accuracy differences measured in the per-iteration tests would amount to differences in convergence rates among the methods if they were allowed to solve problems normally. The question is posed, given that step-size $h$ is heuristically selected by \emph{Peridigm} for the finite difference methods and presumable makes a good choice, do the single-iteration tangent-stiffness accuracy results predict the ranking of the methods in terms of convergence rate for a given study? For brevity, convergence rate was measured for each of the methods in solving just the largest problem of the single-core test series. That problem was solved with each of the methods both on a single core and all sixteen cores of a  node on \emph{Stampede}. As was seen in the per-iteration results, tangent-stiffness accuracy did not vary greatly from test to test, such that any pair of tests from the two series could have been considered reasonably representative. As in the validation problem shown in Section~\ref{subsec:Validation}, would each of the methods return an equal convergence rate so long as $h$ were optimally selected?

The results shown in Table~\ref{tab:ConvergenceStudy2} indicate that this was the case.  The difference in tangent-stiffness accuracy between the methods seen in the per-iteration tests was not sufficient to lead to differences in the convergence rate of the methods.
%
\begin{table}[htb]   
    \centering 
    \caption{Average iterations / load-step} 
    \label{tab:ConvergenceStudy2}   
    \begin{tabular}{c c c c c}
        \toprule 
        Test type & AD & CS & CD & FD\\
        \midrule 
        single-core & 63 & " & " & "\\ 
        multi-core  & 63 & " & " & "\\ 
        \bottomrule 
    \end{tabular} 
\end{table}

The end displacement results shown in Section~\ref{subsec:Validation} indicated that up to a certain precision the solutions produced by each of the methods were identical. It was examined if this was the case for the solutions produced in \emph{Peridigm}. Here each of the solutions produced by each of the methods were compared to the AD produced solutions, under the assumption that the AD produced solutions would be closest to the solution produced by using an 'exact' tangent-stiffness. Table~\ref{tab:PeridigmSolutionAccuracy} shows that the methods did indeed produce differing solutions in terms of nodal force density, although only in decimal places at least $9$ orders of magnitude smaller than the force density value at the node where the difference was detected.
%
\begin{table}[hbp]   
    \centering 
    \caption{Nodal force density maximum difference} 
    \label{tab:PeridigmSolutionAccuracy}   
    \begin{tabular}{c c c c}
        \toprule  & CS & CD & FD\\
        \midrule  Order of magnitude & $\num{1.E-8}$ & $\num{1.E-8}$ & $\num{1.E-5}$ \\ 
        \bottomrule 
    \end{tabular} 
\end{table}

The solution accuracy results do follow the same ranking seen in the tangent-stiffness accuracy results; however, again as in the validation problem the results differ only slightly. Speculatively this degree of error may be important to those running probabilistic simulations such that the propagation of error may be a concern. Also it is conceivable that if this error is a fixed quantity, insensitive to material properties or loading conditions, it may have a greater impact on simulations involving very low force densities such as for small displacements or very compliant materials. In this case it could be recommended that forward difference be avoided. 
%
\section{\todo[]{Re: I.5}Conclusions}
\label{sec:Conc}
%
%Let's leave this section out and see what the reviewer says.  The paper is getting pretty long.
%
%\subsection{For further studies of tangent-stiffness accuracy in Newton's method}
%\label{subsec:FurtherStudies}
%The question is raised; what is the significance of tangent-stiffness estimation accuracy to the convergence of Newton's method?  The per-iteration and even convergence studies shown here indicate the effect of tangent-stiffness accuracy for only a single datum, a very accurate tangent-stiffness estimate, while the validation problem of Section~XX indicates the effect of using a grossly inaccurate tangent-stiffness. The results here show that for very small tangent-stiffness estimate inaccuracies, convergence rate is not affected, while for grossly inaccurate tangent-stiffnesss Newton's method fails.  Additional confounding factors include that a tangent-stiffness is only first order approximation of an often curved energy surface and that a non-naive implementation of Newton's method includes a line search phase that helps in choosing an optimal iterate given a tangent-stiffness.  Is it worthwhile then to investigate the effect of using moderately inaccurate tangent-stiffnesss, produced deliberately or otherwise, on the convergence rate of Newton's method?

%The answer may be no at least unless a more highly non-linear class of problem is selected for study or a controllable method of producing a moderately inaccurate tangent-stiffness is found along with a reason for using it. While it is possible to incorrectly apply finite differencing and CS by choosing an inappropriate step-size, there is no reason that this should be done. The experiments show that step-size $h$ can be properly selected and that the methods are capable of producing accurate tangent-stiffnesss. It should be noted that the model used to perform the simulation could always be substituted for a more highly non-linear one and perhaps this would somewhat increase the importance of tangent-stiffness accuracy. The study presented here failed to pose the question of how wrong a correctly implemented tangent-stiffness calculation scheme could go for plastic or otherwise non-linear material models, here instead non-linearity was imposed by the nonlocal interaction between material points essential to peridynamic models, like the linear model used in this study.
